A Formalization of Typed Lua

### 2015-06-14 doc/papers/paper/paper.tex
DLS paper: some last-minute fixes and additions

\begin{figure*}[t]
$$
\begin{array}{rlr}
F ::= & & \textsc{first-level types:}\\
& \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; F_{1} \cup F_{2} & \textit{union types}\\
& | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
& | \; \{\overline{K:V}\}_{unique|open|fixed|closed} & \textit{table types}\\
& |\; x & \textit{type variables}\\
& | \; \mu x.F & \textit{recursive types}\\
%\multicolumn{3}{c}{}\\
L ::= & & \textsc{{\small literal types:}}\\
& \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
%\multicolumn{3}{c}{}\\
B ::= & & \textsc{{\small base types:}}\\
& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\\\
%\multicolumn{3}{c}{}\\
%\multicolumn{3}{c}{}\\
S ::= & & \textsc{second-level types:}\\
& \;\; P & \textit{tuple types}\\
& | \; S_{1} \sqcup S_{2} & \textit{unions of tuple types}\\
%\multicolumn{3}{c}{}\\
P ::= & & \textsc{{\small tuple types:}}\\
%& \;\; \Void & \textit{void type}\\
& | \; F{*} & \textit{variadic types}\\
& | \; F \times P & \textit{pair types}\\\\
\end{array}
$$
\caption{Type Language}
\label{fig:typelang}
\end{figure*}

\begin{figure*}[t]
$$
\begin{array}{rlr}
T ::= & & \textsc{{\small expression types:}}\\
& \;\; V & \\
& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
E ::= & & \textsc{{\small expression list types:}}\\
& \;\; T{*} & \\
& | \; T \times E & \\
K ::= & & \textsc{{\small key types:}}\\
& \;\; L \; | \; B \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
V ::= & & \textsc{{\small value types:}}\\
& \;\; F \; | \; \Const \; F &\\
R ::= & & \textsc{\small{filter result type:}}\\
& \;\; \mathbf{void} & \textit{void type}\\
& |\; F
\end{array}
$$
\caption{Special Types}
\label{fig:special}
\end{figure*}

@@

Literal types represent the type of literal boolean, number
and string values. Their main application is the formation in record and tuple types.

@@

Table types represent some of the forms that Lua
tables can take. What a particular table type denotes
depends on its {\em tag}: the {\em unique}, {\em open}
and {\em fixed} tags all denote tables where it is
guaranteed that every key belongs to one of the {\em key
types} $K_i$. We restrict key types to literal types,
base types, and the top type to make the system simpler
for the user, as our survey of Lua programs showed that
this restriction reflects the way Lua tables are commonly used.

@@

\begin{figure}[t]
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-TableFCtoC}\\
\dfrac{\begin{array}{c}
	\forall j \; \exists i \;\;\;
	\senv \vdash K_{i} \subtype K_{j}' \;\
	\senv \vdash K_{j}' \subtype K_{i} \;
	\senv \vdash V_{i} \subtype_{c} V_{j}'
	\end{array}}
{\begin{array}{c}
	\senv \vdash \{\overline{K:V}\}_{fixed|closed} \subtype
	\{\overline{K^\prime:V^\prime}\}_{closed}
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableUtoC}\\
\dfrac{\begin{array}{c}
	\forall i \; \forall j \;\;\;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{u} V_{j}' \\
	\forall j  \; \nexists i \;\;\;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\senv \vdash \{\overline{K:V}\}_{unique} \subtype
	\{\overline{K^\prime:V^\prime}\}_{closed}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableOtoC}\\
\dfrac{\begin{array}{c}
	\forall i \; \forall j \;\;\;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{c} V_{j}' \\
	\forall j \; \nexists i \;\;\;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\senv \vdash \{\overline{K:V}\}_{open} \subtype
	\{\overline{K^\prime:V^\prime}\}_{closed}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableUtoUOF}\\
\dfrac{\begin{array}{c}
	\forall i \; \exists j \;\;\;
	\senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{u} V_{j}' \\
	\forall j  \; \nexists i \;\;\;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\begin{array}{c}
	\senv \vdash \{\overline{K:V}\}_{unique} \subtype
	\{\overline{K^\prime:V^\prime}\}_{unique|open|fixed}
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableOtoOF}\\
\dfrac{\begin{array}{c}
	\forall i \; \exists j \;\;\;
	\senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{c} V_{j}' \\
	\forall j \; \nexists i \;\;\;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\begin{array}{c}
	\senv \vdash \{\overline{K:V}\}_{open} \subtype
	\{\overline{K^\prime:V^\prime}\}_{open|fixed}
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableFtoF}\\
\dfrac{\begin{array}{c}
	\forall j \; \exists i \;\;\;
	\senv \vdash K_{i} \subtype K_{j}' \;
	\senv \vdash K_{j}' \subtype K_{i} \;
	\senv \vdash V_{i} \subtype_{c} V_{j}'\\
	\forall i \; \exists j \;\;\;
	\senv \vdash K_{i} \subtype K_{j}' \;
	\senv \vdash K_{j}' \subtype K_{i} \;
	\senv \vdash V_{i} \subtype_{c} V_{j}'\\
	\end{array}}
{\senv \vdash \{\overline{K:V}\}_{fixed} \subtype
	\{\overline{K^\prime:V^\prime}\}_{fixed}}
\end{array}
\end{array}
\]
\caption{Subtyping for table types}
\label{fig:tablesub}
\end{figure}

@@

\begin{figure*}[t]
	$$
	\begin{array}{rlr}
	s ::= & & \textsc{statements:}\\
	& \;\; \mathbf{skip} & \textit{skip}\\
	& | \; s_{1} \; ; \; s_{2} & \textit{sequence}\\
	& | \; \overline{l} = el & \textit{multiple assignment}\\
	& | \; \mathbf{while} \; e \; \mathbf{do} \; s \;
	| \; \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2} & \textit{control flow}\\
	& | \; \mathbf{local} \; \overline{id{:}F} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
	& | \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
	& | \; \mathbf{rec} \; id{:}F = f \; \mathbf{in} \; s & \textit{recursive function} \\
	& | \; \mathbf{return} \; el & \textit{return} \\
	& | \; \lfloor a \rfloor_{0} & \textit{application with no results}\\
	& | \; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \textit{method declaration}\\
	e ::= & & \textsc{expressions:}\\
	& \;\; \mathbf{nil} & \textit{nil}\\
	& | \; k & \textit{other literals}\\
	& | \; id & \textit{variable access}\\
	& | \; e_{1}[e_{2}] & \textit{table access}\\
	& | \; {<}F{>} \; id & \textit{type coercion}\\
	& | \; f & \textit{function declaration}\\
	& | \; \{ \; \overline{[e_{1}] = e_{2}} \; \} \;
	| \; \{ \; \overline{[e_{1}] = e_{2}},me \; \} & \textit{table constructor}\\
	& | \; e_{1} + e_{2} \;
	| \; e_{1} \; {..} \; e_{2} \;
	| \; e_{1} == e_{2} \;
	| \; e_{1} < e_{2} & \textit{binary operations}\\
	& | \; e_{1} \;\&\; e_{2} \;
	| \; e_{1} \; \mathbf{and} \; e_{2} \;
	| \; e_{1} \; \mathbf{or} \; e_{2} & \textit{binary operations}\\
	& | \; \mathbf{not} \; e \;
	| \; \# \; e & \textit{unary operations} \\
	& | \; \lfloor me \rfloor_{1} & \textit{expressions with one result}\\
	l ::= & & \textsc{left-hand values:}\\
	& \;\; id_{l} & \textit{variable assignment}\\
	& | \; e_{1}[e_{2}] & \textit{table assignment}\\
	& | \; id[k] \; {<}V{>} & \textit{type coercion}\\
	k ::= & & \textsc{literal constants:}\\
	& \;\; \mathbf{false} \; | \;
	\mathbf{true} \; | \;
	{\it int} \; | \;
	{\it float} \; | \;
	{\it string} & \\
	el ::= & & \textsc{expression lists:}\\
	& \;\; \overline{e} \; | \;
	\overline{e}, me & \\
	me ::= & & \textsc{multiple results:}\\
	& \;\; a & \textit{application}\\
	& | \; {...} & \textit{vararg expression}\\
	a ::= & & \textsc{applications:}\\
	& \;\; e(el) & \textit{function application}\\
	& | \; e{:}n(el) & \textit{method application}\\
	f ::= & & \textsc{function declarations:}\\
	& \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
	pl ::= & & \textsc{parameter lists:}\\
	& \;\; \overline{id{:}F} \; | \;
	\overline{id{:}F},{...}{:}F & \\
	\end{array}
	$$
	\caption{Abstract Syntax}
	\label{fig:syntax}
\end{figure*}

@@

\begin{figure*}[t]
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-Constructor}\\
\dfrac{\begin{array}{c}
\overline{\env_1, \penv \vdash e_2: F, \env_v}
	\;\;\;
	\overline{\env_v, \penv \vdash e_1: K, \env_f}
	\\
	T = \{\overline{K{:}fix(F)}\}_{unique} \;\;\;
	\env_{2} = merge(\overline{\env_{f}}) \;\;\; wf(T)
	\end{array}}
{\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{2}}
\end{array}
\begin{array}{c}
\mylabel{T-NewFieldUnique}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \{ \overline{K:V} \}_{unique}\\
	\env_{1}, \penv \vdash k:K^{new}, \env_{2} \;\;\;
	\overline{K^{new} \not\lesssim K} \;\;\;
	V^{new} = fix(V)
	\end{array}}
{\begin{array}{l}
	\env_{1}, \penv \vdash id[k] {<}V{>}:V^{new},
	\env_{2}[id \mapsto \{ \overline{K{:}V}, K^{new}{:}V^{new}\}_{unique}]
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{T-IdRead}\\
\dfrac{\env(id) = F}
{\env, \penv \vdash id:close(F), \env[id \mapsto open(F)]}
\end{array}
\begin{array}{c}
\mylabel{T-IdWrite}\\
\dfrac{\env(id) = F}
{\env, \penv \vdash id_{l}:F, \env}
\end{array}
\begin{array}{c}
\mylabel{T-IndexId}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \{\overline{K:V}\} \;\;\;
	\env_{1}, \penv \vdash e_{2}:K^e, \env_{2} \;\;\;
	 K^e \lesssim K_{i}
	\end{array}}
{\env_{1}, \penv \vdash id[e_{2}]:V_{i}, \env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-IndexExp}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash e_{1}:\{\overline{K:V}\}, \env_{2}\\
	\env_{2}, \penv \vdash e_{2}:K^e, \env_{3} \;\;\;
	 K^e \lesssim K_{i}
	\end{array}}
{\env_{1}, \penv \vdash e_{1}[e_{2}]:V_{i}, \env_{3}}
\end{array}
\begin{array}{c}
\mylabel{T-CoerceClosed}\\
\dfrac{\env_{1}(id) \subtype F \;\;\; tag(F,closed)}
{\env, \penv \vdash {<}F{>} \; id:F, \env[id \mapsto reopen(F)]}
\end{array}
\begin{array}{c}
\mylabel{T-CoerceFixed}\\
\dfrac{\env_{1}(id) \subtype F \;\;\; tag(F,fixed)}
{\env, \penv \vdash {<}F{>} \; id:F, \env[id \mapsto F]}
\end{array}\\
\begin{array}{c}
\mylabel{T-Function}\\
\dfrac{\begin{array}{c}
	closeall(\env_{1})[\overline{id \mapsto F}], \penv[\ret \mapsto S] \vdash s, \env_{2}\\
	\env_{3} = openset(\env_{1}, frv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s)) \\
	\env_{4} = closeset(\env_{3}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}F}){:}S \; s:F_1 \times \ldots F_n \times \Void \rightarrow S, \env_{3}}
\end{array}
\end{array}
\]
\caption{Rules for Table Types}
\label{fig:tablerules}
\end{figure*}

@@

Well-formed table types also do not allow \emph{unique} and
\emph{open} table types to appear in the type of the values.
Our type system currently cannot
track aliases of table fields, so this restriction is
necessary. This limitation of the type system is the reason the {\sc T-Constructor} rule uses the auxiliary function $fix$ to change any {\em open} and {\em unique} table types used in the field initializer to {\em fixed}. As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
has type $\{``x":1, ``y":\{``z":2\}_{fixed}\}_{unique}$.

@@

Open tables have a rule {\sc T-NewFieldOpen} that is not shown
in Figure~\ref{fig:tablerules} due to brevity. The only thing that
changes from one rule to the other is the tag in the type of $id$.

@@

Consider the following example, where we make an alias
of a {\em unique} table:
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{closed} = a \; \mathbf{in}$}\\
& & $\;a[``x"] \; {<}\String{>} = ``foo";$
\end{tabular}
\end{center}

@@

Another use of the identifier that does not create an
alias is when indexing the table. Rule {\sc T-IndexId}
covers this case. For comparison, rule {\sc T-Index}
applies for indexing expressions where the expression
denoting the table is not an identifier.

@@

@@

proj(F{*}, i) & = & F \cup \Nil \\

@@

\]

>>>>

\begin{figure*}[t]
$$
\begin{array}{rlr}
F ::= & & \textsc{first-level types:}\\
& \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; F_{1} \cup F_{2} & \textit{union types}\\
& | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
& | \; \{\overline{F:V}\}_{unique|open|fixed|closed} & \textit{table types}\\
& |\; x & \textit{type variables}\\
& | \; \mu x.F & \textit{recursive types}\\
%\multicolumn{3}{c}{}\\
L ::= & & \textsc{{\small literal types:}}\\
& \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
%\multicolumn{3}{c}{}\\
B ::= & & \textsc{{\small base types:}}\\
& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\\\
%\multicolumn{3}{c}{}\\
%\multicolumn{3}{c}{}\\
S ::= & & \textsc{second-level types:}\\
& \;\; P & \textit{tuple types}\\
& | \; S_{1} \sqcup S_{2} & \textit{unions of tuple types}\\
%\multicolumn{3}{c}{}\\
P ::= & & \textsc{{\small tuple types:}}\\
%& \;\; \Void & \textit{void type}\\
& | \; F{*} & \textit{variadic types}\\
& | \; F \times P & \textit{pair types}\\\\
\end{array}
$$
\caption{Type Language}
\label{fig:typelang}
\end{figure*}

\begin{figure*}[t]
$$
\begin{array}{rlr}
T ::= & & \textsc{{\small expression types:}}\\
& \;\; V & \\
& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
E ::= & & \textsc{{\small expression list types:}}\\
& \;\; T{*} & \\
& | \; T \times E & \\
%K ::= & & \textsc{{\small key types:}}\\
%& \;\; L \; | \; B \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
V ::= & & \textsc{{\small value types:}}\\
& \;\; F \; | \; \Const \; F &\\
R ::= & & \textsc{\small{filter result type:}}\\
& \;\; \mathbf{void} & \textit{void type}\\
& |\; F
\end{array}
$$
\caption{Special Types}
\label{fig:special}
\end{figure*}

@@

Literal types represent the type of literal boolean, number
and string values. Having just one value each, their main application is in constructing table types with fields where it is guaranteed that the value of that field will not be {\tt nil}.

@@

Table types represent some of the forms that Lua
tables can take. What a particular table type denotes
depends on its {\em tag}: the {\em unique}, {\em open}
and {\em fixed} tags all denote tables where it is
guaranteed that every key belongs to one of the {\em key
types} $F_i$. 
%We restrict key types to literal types,
%base types, and the top type to make the system simpler
%for the user, as our survey of Lua programs showed that
%this restriction reflects the way Lua tables are commonly used.

@@

\begin{figure}[t]
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-TableFCtoC}\\
\dfrac{\begin{array}{c}
	\forall j \; \exists i \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \;\
	\senv \vdash F_{j}' \subtype F_{i} \;
	\senv \vdash V_{i} \subtype_{c} V_{j}'
	\end{array}}
{\begin{array}{c}
	\senv \vdash \{\overline{F:V}\}_{fixed|closed} \subtype
	\{\overline{F^\prime:V^\prime}\}_{closed}
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableUtoC}\\
\dfrac{\begin{array}{c}
	\forall i \; \forall j \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash V_{i} \subtype_{u} V_{j}' \\
	\forall j  \; \nexists i \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\senv \vdash \{\overline{F:V}\}_{unique} \subtype
	\{\overline{F^\prime:V^\prime}\}_{closed}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableOtoC}\\
\dfrac{\begin{array}{c}
	\forall i \; \forall j \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash V_{i} \subtype_{c} V_{j}' \\
	\forall j \; \nexists i \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\senv \vdash \{\overline{F:V}\}_{open} \subtype
	\{\overline{F^\prime:V^\prime}\}_{closed}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableUtoUOF}\\
\dfrac{\begin{array}{c}
	\forall i \; \exists j \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \land \senv \vdash V_{i} \subtype_{u} V_{j}' \\
	\forall j  \; \nexists i \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\begin{array}{c}
	\senv \vdash \{\overline{F:V}\}_{unique} \subtype
	\{\overline{F^\prime:V^\prime}\}_{unique|open|fixed}
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableOtoOF}\\
\dfrac{\begin{array}{c}
	\forall i \; \exists j \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \land \senv \vdash V_{i} \subtype_{c} V_{j}' \\
	\forall j \; \nexists i \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\begin{array}{c}
	\senv \vdash \{\overline{F:V}\}_{open} \subtype
	\{\overline{F^\prime:V^\prime}\}_{open|fixed}
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableFtoF}\\
\dfrac{\begin{array}{c}
	\forall j \; \exists i \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \;
	\senv \vdash F_{j}' \subtype F_{i} \;
	\senv \vdash V_{i} \subtype_{c} V_{j}'\\
	\forall i \; \exists j \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \;
	\senv \vdash F_{j}' \subtype F_{i} \;
	\senv \vdash V_{i} \subtype_{c} V_{j}'\\
	\end{array}}
{\senv \vdash \{\overline{F:V}\}_{fixed} \subtype
	\{\overline{F^\prime:V^\prime}\}_{fixed}}
\end{array}
\end{array}
\]
\caption{Subtyping for table types}
\label{fig:tablesub}
\end{figure}

@@

\begin{figure*}[t]
	$$
	\begin{array}{rlr}
	s ::= & & \textsc{statements:}\\
	& \;\; \mathbf{skip} & \textit{skip}\\
	& | \; s_{1} \; ; \; s_{2} & \textit{sequence}\\
	& | \; \overline{l} = el & \textit{multiple assignment}\\
	& | \; \mathbf{while} \; e \; \mathbf{do} \; s \;
	| \; \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2} & \textit{control flow}\\
	& | \; \mathbf{local} \; \overline{id{:}F} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
	& | \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
	& | \; \mathbf{rec} \; id{:}F = f \; \mathbf{in} \; s & \textit{recursive function} \\
	& | \; \mathbf{return} \; el & \textit{return} \\
	& | \; \lfloor a \rfloor_{0} & \textit{application with no results}\\
	& | \; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \textit{method declaration}\\
	e ::= & & \textsc{expressions:}\\
	& \;\; \mathbf{nil} & \textit{nil}\\
	& | \; k & \textit{other literals}\\
	& | \; id & \textit{variable access}\\
	& | \; e_{1}[e_{2}] & \textit{table access}\\
	& | \; {<}F{>} \; id & \textit{type coercion}\\
	& | \; f & \textit{function declaration}\\
	& | \; \{ \; \overline{[e_{1}] = e_{2}} \; \} \;
	| \; \{ \; \overline{[e_{1}] = e_{2}},me \; \} & \textit{table constructor}\\
	& | \; e_{1} + e_{2} \;
	| \; e_{1} \; {..} \; e_{2} \;
	| \; e_{1} == e_{2} \;
	| \; e_{1} < e_{2} & \textit{binary operations}\\
	& | \; e_{1} \;\&\; e_{2} \;
	| \; e_{1} \; \mathbf{and} \; e_{2} \;
	| \; e_{1} \; \mathbf{or} \; e_{2} & \textit{binary operations}\\
	& | \; \mathbf{not} \; e \;
	| \; \# \; e & \textit{unary operations} \\
	& | \; \lfloor me \rfloor_{1} & \textit{expressions with one result}\\
	l ::= & & \textsc{left-hand values:}\\
	& \;\; id_{l} & \textit{variable assignment}\\
	& | \; e_{1}[e_{2}]_{l} & \textit{table assignment}\\
	& | \; id[k] \; {<}V{>} & \textit{type coercion}\\
	k ::= & & \textsc{literal constants:}\\
	& \;\; \mathbf{false} \; | \;
	\mathbf{true} \; | \;
	{\it int} \; | \;
	{\it float} \; | \;
	{\it string} & \\
	el ::= & & \textsc{expression lists:}\\
	& \;\; \overline{e} \; | \;
	\overline{e}, me & \\
	me ::= & & \textsc{multiple results:}\\
	& \;\; a & \textit{application}\\
	& | \; {...} & \textit{vararg expression}\\
	a ::= & & \textsc{applications:}\\
	& \;\; e(el) & \textit{function application}\\
	& | \; e{:}n(el) & \textit{method application}\\
	f ::= & & \textsc{function declarations:}\\
	& \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
	pl ::= & & \textsc{parameter lists:}\\
	& \;\; \overline{id{:}F} \; | \;
	\overline{id{:}F},{...}{:}F & \\
	\end{array}
	$$
	\caption{Abstract Syntax}
	\label{fig:syntax}
\end{figure*}

@@

\begin{figure*}[t]
{\small
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-Constructor}\\
\dfrac{\begin{array}{c}
	\overline{\env_v, \penv \vdash e_1: F_1, \env_f}
	\;\;\;
\overline{\env_1, \penv \vdash e_2: F_2, \env_v}	\\
	T = \{\overline{F_1{:}vt(F_1,F_2)}\}_{unique} \;\;\;
	\env_{2} = merge(\overline{\env_{f}}) \;\;\; wf(T)
	\end{array}}
{\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{2}}
\end{array}
\begin{array}{c}
\mylabel{T-NewFieldUnique}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \{ \overline{F:V} \}_{unique}\\
	\env_{1}, \penv \vdash k:F^{new}, \env_{2} \;\;\;
	\overline{F^{new} \not\lesssim F} \;\;\;
	V^{new} = vt(F^{new}, V)
	\end{array}}
{\begin{array}{l}
	\env_{1}, \penv \vdash id[k] {<}V{>}:V^{new},
	\env_{2}[id \mapsto \{ \overline{F{:}V}, F^{new}{:}V^{new}\}_{unique}]
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{T-IdRead}\\
\dfrac{\env(id) = F}
{\env, \penv \vdash id:close(F), \env[id \mapsto open(F)]}
\end{array}
\begin{array}{c}
\mylabel{T-IdWrite}\\
\dfrac{\env(id) = F}
{\env, \penv \vdash id_{l}:F, \env}
\end{array}
\begin{array}{c}
\mylabel{T-IndexIdRead}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \{\overline{F:V}\} \;\;\;
	\env_{1}, \penv \vdash e_{2}:F^e, \env_{2} \;\;\;
	 F^e \lesssim F_{i}
	\end{array}}
{\env_{1}, \penv \vdash id[e_{2}]:rconst(V_{i}), \env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-IndexExpRead}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash e_{1}:\{\overline{F:V}\}, \env_{2}\\
	\env_{2}, \penv \vdash e_{2}:F^e, \env_{3} \;\;\;
	 F^e \lesssim F_{i}
	\end{array}}
{\env_{1}, \penv \vdash e_{1}[e_{2}]:rconst(V_{i}), \env_{3}}
\end{array}
\begin{array}{c}
\mylabel{T-CoerceClosed}\\
\dfrac{\env_{1}(id) \subtype F \;\;\; tag(F,closed)}
{\env, \penv \vdash {<}F{>} \; id:F, \env[id \mapsto reopen(F)]}
\end{array}
\begin{array}{c}
\mylabel{T-CoerceFixed}\\
\dfrac{\env_{1}(id) \subtype F \;\;\; tag(F,fixed)}
{\env, \penv \vdash {<}F{>} \; id:F, \env[id \mapsto F]}
\end{array}\\\\
\begin{array}{c}
\mylabel{T-Function}\\
\dfrac{\begin{array}{c}
	closeall(\env_{1})[\overline{id \mapsto F}], \penv[\ret \mapsto S] \vdash s, \env_{2}\\
	\env_{3} = openset(\env_{1}, frv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s)) \\
	\env_{4} = closeset(\env_{3}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}F}){:}S \; s:F_1 \times \ldots F_n \times \Void \rightarrow S, \env_{3}}
\end{array}
\begin{array}{c}
\mylabel{T-Local}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash el:E, \env_{2} \\
	E \lesssim F_1 \times \ldots \times F_n \times \Value{*} \\
	\env_{2}[\overline{id \mapsto F}], \penv \vdash s, \env_{3}
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{local} \; \overline{id{:}F} = el \; \mathbf{in} \; s, (\env_{3} - \{\overline{id}\})[\overline{id \mapsto \env_2(id)}]}
\end{array}
\end{array}
\]
}
\caption{Typing Rules}
\label{fig:tablerules}
\end{figure*}

@@

Well-formed table types also do not allow \emph{unique} and
\emph{open} table types to appear in the type of the values.
Our type system currently cannot
track aliases of table fields, so this restriction is
necessary. This limitation of the type system is the reason the {\sc T-Constructor} rule uses the auxiliary function $fix$
(in the definition of $vt$)
 to change any {\em open} and {\em unique} table types used in the field initializer to {\em fixed}. As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
has type $\{``x":1, ``y":\{``z":2\}_{fixed}\}_{unique}$.

@@

Open tables have a rule {\sc T-NewFieldOpen} that is not shown
in Figure~\ref{fig:tablerules} due to brevity. The only thing that
changes from one rule to the other is the tag in the type of $id$.
There are analogous rules for adding methods to a {\em unique}
and {\em open tables} as a side-effect of typechecking a method declaration.

@@

Consider the following example, where we make an alias
of a {\em unique} table:
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{closed} = a \; \mathbf{in}\;\;a[``x"] \; {<}\String{>} = ``foo";$}
\end{tabular}
\end{center}

@@

Another use of the identifier that does not create an
alias is when indexing the table. Rule {\sc T-IndexIdRead}
covers this case (function $rconst$ just strips the {\bf const} qualifier from the value type, if present). For comparison, rule {\sc T-IndexExpRead}
applies for indexing expressions where the expression
denoting the table is not an identifier. The rules for
indexing expressions in l-value position are similar,
except to make sure that the field is not {\bf const} .

@@

{\small

@@

proj(F{*}, i) & = & nil(F) \\

@@

\]}
